Response to RC1 of ``When does vapor pressure deficit drive or reduce evapotranspiration?''

### 2018-12-20 doc/paper/hess-revision/review-2.tex
response 2: what I sent to pierre changjie for edits

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{xcolor}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
\usepackage{color}
\pagestyle{plain}

@@

Also please note our response to RC1, as some of the discussion there
is relelvant to this response; particularly discussion on PFT-scale
analysis \url{https://editor.copernicus.org/index.php/hess-2018-553-AC1.pdf?_mdl=msover_md&_jrl=13&_lcm=oc108lcm109w&_acm=get_comm_file&_ms=72556&c=153745&salt=17166479111051651323}.

\section{Major comment (i)}

One issue with analyses attempting to diagnose the ET response to VPD
with pure data (i.e. using binned boxplots, as you suggest), is that
ET varies strongly with both aerodynamic resistance and
radiation. Simply binning by VPD or by site includes too many
confounding factors to really diagnose what the attirbutable response
is to VPD. This issue is really what motivated our analysis and makes
it unique; by building an analytical framework based on confirmed
results in the literatre (uWUE, Medlyn model, Penman-Monteith), we are
able to formally deduce what the actual ET response is, with other
environmental factors held fixed. This ``other environmental factors
held fixed'' component is what is, in our mind, nearly impossible to
deduce from pure data, because the environment is always changing and
direct analogues are either impossible to find or result in too small
of a sample size to form a meaningful conclusion. However, while we
cannot assess our formulation of the derivative with respect to VPD,
we can assess the actual model of ET, which includes all
approximations we introudced (specifically uWUE). As a part of our
analysis test suite we generated a figure of this comparison and
reproduce it here (Figure \ref{rmse}). Apologies this is not a
publication quality figure, but we wanted to get this response out
sooner rather than later to stimulate discussion, especially as we are
already behind because of the AGU Fall Meeting. This comparison is
fair in the sense that by using uWUE we introduce an extra free
parameter, so in the original PM model we introduce an extra free
parameter modifying g$_s$ and fit it also for each given PFT. The
introduction of the uWUE approximation does degrade the quality of the
ET model relative to the original PM model. However, we expect this,
as using uWUE is a simplification over using GPP directly, and may
break down in extreme, limiting cases like VPD $\to$ 0. As
a purpose-built model for assessing the ET response, we think the uWUE
approach is justified. Essentially we are saying that to the degree we
trust arithmetic, uWUE, Penman-Monteith, and the Medlyn stomatal
conductance model, our evaluation of the derivative is robust.

\begin{figure}[h]
  \centering \includegraphics{./rmse_bias.png}
  \caption{Mean bias and root mean squared error (RMSE) for three
    different ET estimates, relative to FLUXNET-2015 observations:
    original (original Penman-Monteith formulation, e.g. Equation 1 in
    the manuscript, with Equation 4 providing g$_s$; uWUE (Equation 8
    in the manuscript); iWUE (as in Equation 8, but if we had used
    iWUE to remove GPP dependence rather than uWUE).}
  \label{rmse}
\end{figure}

However, the discussion so far assumes that the variables in our model
(Equation 8) do not change directly with VPD and we are free to
evaluate the derivative. You bring up a very good example that we do
not explictly consider: a change in VPD will induce a change in
surface temperature, which will alter the net radiation term. We will
definitely add language addressing this oversight. Ideally, we would
incorporate that effect directly in Equation 8. We briefly tried to do
this in time for this response, but difficulties deriving a diagnostic
equation for temperature sabatoged the effort, and we do not want to
delay the response further. We will continue to work on this for the
final response phase. In the meantime, I think we can make some
hand-wavy arguments in the interest of discussion as to what the
effect would be. In the case where the aerodynamic and physiological
ET response (which the manuscript does consider) increases with VPD,
we would expect a decrease in surface temperature. This would induce
an increase in net radiation, as OLR and ground heat flux
decrease. This increase in net radation would induce a further
increase in ET response. Granted, the net radiation perturbation would
damp the decreased temperature response to the
aerodynamic/physiological terms, but all of the signs of the change
would stay the same. The converse is true for a decrease in ET with
respect to VPD: surface temperature increases, OLR and ground heat
flux increase, net radiation decreases, ET decreases. In this way, we
believe the effects of VPD on radiation do not change the nature of
the sign of the response, but could amplify the magnitude. Comments on
this are incouraged, and we will work to include a more quantitative or
rigorous analysis in the final response. Sometimes it is easy to trick
onself with these types of qualitative arguments, particularly with
complicated systems, so hopefully we have not presented faulty logic
in our rush to make this response public.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{xcolor}
\usepackage{hyperref}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
\usepackage{color}
\pagestyle{plain}

\graphicspath{{../../../etc/plots/climate_et/test_plots/}}

@@

Also please note our response to RC1, as some of the discussion there
is relevant to this response; particularly discussion on PFT-scale
analysis and variability of g$_1$ and uWUE
(\href{https://editor.copernicus.org/index.php/hess-2018-553-AC1.pdf?_mdl=msover_md&_jrl=13&_lcm=oc108lcm109w&_acm=get_comm_file&_ms=72556&c=153745&salt=17166479111051651323}{RC1
Response}).

\section{Major comment (i)}
\label{i}
One issue with analyses attempting to diagnose the ET response to VPD
with pure data (e.g. using binned boxplots, as you suggest), is that
ET varies strongly with both aerodynamic resistance and
radiation. Simply binning by VPD or by site includes too many
confounding factors to really diagnose what the attributable response
is to VPD. This issue is really what motivated our analysis and makes
it unique; by building an analytical framework based on confirmed
results in the literature (uWUE, Medlyn model, Penman-Monteith [PM]), we are
able to formally deduce what the actual ET response is, with other
environmental factors held fixed. This ``other environmental factors
held fixed'' component is, in our mind, nearly impossible to
deduce from pure data, because the environment is always changing and
direct analogues are either impossible to find or result in too small
of a sample size to form a meaningful conclusion. However, while we
cannot assess our formulation of the derivative with respect to VPD,
we can assess the actual model of ET, which includes all
approximations we introduced (specifically uWUE). As a part of our
analysis test suite we generated a figure of this comparison and
reproduce it here (Figure \ref{rmse}). Apologies this is not a
publication quality figure, but we wanted to get this response out
sooner rather than later to stimulate discussion, especially as we are
already behind because of the AGU Fall Meeting. This comparison is
fair in the sense that by using uWUE we introduce an extra free
parameter, so in the original PM model we introduce an extra free
parameter modifying g$_s$ and fit it also for each given PFT. The
introduction of the uWUE approximation does degrade the quality of the
ET model relative to the original PM model. However, we expect this,
as using uWUE is a simplification over using GPP directly, and may
break down in extreme, limiting cases like VPD $\to$ 0. As
a purpose-built model for assessing the ET response, we think the uWUE
approach is justified. Essentially we are saying that to the degree we
trust arithmetic, uWUE, Penman-Monteith, and the Medlyn stomatal
conductance model, our evaluation of the derivative is robust.

\begin{figure}
  \centering \includegraphics[width=0.75\textwidth]{rmse_bias.png}
  \caption{Mean bias (absolute value) and root mean squared error
    (RMSE) for three different ET estimates, relative to FLUXNET-2015
    observations: original (original Penman-Monteith formulation,
    e.g. Equation 1 in the manuscript, with Equation 4 providing
    g$_s$); uWUE (Equation 8 in the manuscript); iWUE (as in Equation
    8, but if we had used iWUE to remove GPP dependence rather than
    uWUE).}
  \label{rmse}
\end{figure}

However, the discussion so far assumes that the variables in our model
(Equation 8) do not change directly with VPD and we are free to
evaluate the derivative. You bring up a very good example that we did
not consider: a change in VPD will induce a change in net radiation
through surface temperature that PM does not consider. We will
definitely add language addressing this oversight. Ideally, we would
incorporate that effect directly in Equation 8. We briefly tried to do
this in time for this response, but difficulties deriving a diagnostic
equation for temperature sabotaged the effort, and we do not want to
delay the response further. We will continue to work on this for the
final response phase. In the meantime, we think we can make some
logic-based arguments in the interest of discussion as to what the
effect would be. In the case where the aerodynamic and physiological
ET response (which the manuscript does consider) increases with VPD,
we would expect a decrease in surface temperature. This would induce
an increase in net radiation, as OLR and ground heat flux
decrease. This increase in net radiation would induce a further
increase in ET response. Granted, the net radiation perturbation would
damp the decreased temperature response to the
aerodynamic/physiological terms, but all of the signs of the change
would stay the same. The converse is true for a decrease in ET with
respect to VPD: surface temperature increases, OLR and ground heat
flux increase, net radiation decreases, ET decreases. In this way, we
believe the effects of VPD on radiation do not change the nature of
the sign of the response, but could amplify the magnitude. Comments on
this are encouraged, and we will work to include a more quantitative
and rigorous analysis in the final response. Sometimes it is easy to
trick oneself with these types of qualitative arguments, particularly
with complicated systems, so hopefully we have not presented faulty
logic in our rush to make this response public.

\section{Major comment (ii)}
\label{ii}
Thank you for having an open mind about the concave up ET-VPD curve
result and providing clear explanations of concerns about this result
based on literature. Given the apparent controversy, we will
definitely add more language connecting previous results, particularly
at the leaf scale, with the presented results. Here we will informally
elaborate in the interest of discussion.

One of the primary differences between ecosystem scale assessments of
ET (specifically using Penman Monteith) and experimental and modeling
results of leaf scale response to VPD is that by taking an energy
balance approach Penman Monteith accounts for changes in $e_{sat}(T)$
at the surface induced by ET. To demonstrate this, compare Figure
\ref{leaf} and Figure \ref{PM} (``Original Penman Monteith''
Curve). In Figure \ref{leaf} we apply a conceptual leaf scale
approach to calculate ecosystem ET without the energy balance
consideration introduced by PM. In this figure, the curve shows a
concave downward shape, in direct analogy to leaf scale T. Once the
effects of evaporation's thermodynamics are included in the energy
balance, the curve's shape is no longer concave downward (Figure
\ref{PM}, ``Original Penman Monteith'' curve). How these conceptual
results translate to leaf scale experiments depends on the
experimental setup \citep[e.g.,][]{Rawson1977, Turner1984, Mott2013};
however generally chamber-based leaf scale experiments do not preserve
the energy balance relationships we expect for a surface in a
natural environment (sometimes intentionally by design). Additionally,
we would like to note that a concave up result is not necessarily
inconsistent with the statement ``transpiration at the leaf scale is
mostly positive'' with VPD. Especially for plants in well watered
environments like crops, the manuscript's ecosystem scale shape
(Figure \ref{PM}) would not be out of place on the figures in
\cite{Rawson1977} and \cite{Mott2007} for the range of VPD considered
(Figure \ref{PM} compared to Fig. 2 in \cite{Rawson1977}, also
Figure 7 in \cite{Mott2007}). However, the change in relative
magnitudes are not similar, and this effect can be traced partially to
the effect of decreasing surface $e_{sat}(T)$ in response to
increasing evapotranspiration. While not directly considered here (or
in the manuscript) we also think it is important to consider how plant
physiology in the natural environment could vary from plants grown in
well-watered conditions in experiments. It is not inconceivable that
phenotypic variability and adaptation in response to water limitation
could result in plants with different responses to those grown in a
lab. We presented crops in our figures because they are the most likely
to well watered and analogous with laboratory results; however for
other PFTs the concavity of the VPD-ET curves are more pronounced at
higher VPD as compared to Figure \ref{PM}.

\begin{figure}
  \centering \includegraphics[width=1.0\textwidth]{CRO_et_vpd_curve_leaf.png}
  \caption{The conceptual relationship between ET with VPD using a
    model analogous to leaf scale models and experiments, evaluated
    with the study's median ecosystem scale resistances for the crop
    PFT. This estimation of ET is by definition not physically
    representative, but the figure is intended as a conceptual
    description of how leaf-scale theory's ET-VPD relationship deviates
    from energy balance ecosystem theory (Fig. \ref{PM}). This figure is
    demonstrative, and not intended to be of publication quality.}
  \label{leaf}
\end{figure}

\begin{figure}
  \centering \includegraphics[width=1.0\textwidth]{CRO_et_vpd_curve_with_pm.png}
  \caption{The conceptual relationship between ET and VPD using the
    uWUE derived version of Penman Monteith (Equation 8 in the manuscript)
    and the original Penman Monteith formulation (Equation 1 in the
    manuscript), evaluated at the study's median conditions for the
    crop PFT. uWUE Penman Monteith includes the effects of VPD on
    photosynthesis. This figure is demonstrative, and not intended to be of
    publication quality.}
  \label{PM}
\end{figure}

The previous paragraph discussed the change in frame of reference
induced by using PM (and energy balance) to assess ecosystem scale ET
rather than directly applying leaf scale logic in controlled
experiments. However, our uWUE-based PM framework additionally
accounts for changes in photosynthesis induced by VPD (Figure
\ref{PM}). Discussion in \cite{Damour2010} (``The issue of
co-regulation of g$_s$ and A$_{net}$'') provides some interesting
background both on the general weaknesses of current stomatal theory,
and specifically on the possible importance of resolving the effects
of water stress on $A_{net}$ in stomatal conductance. Given that the
current state of the art in physically-derived theory has not
converged on the proper way to account for changes in $A_{net}$ and
$g_s$ \citep{Damour2010}, we believe using the validated uWUE
empiracle results \citep[e.g. Figure 3,][]{Zhou_2014} is a good
approach to develop an estimate of ET response to VPD accounting for
changes in photosynthesis. This method of using an empirical result in
the absence of a complete physical theory has been used often in
science (e.g. turbulence) to increase our knowlege about a given
systems' behavior, and we think it is well motivated here. However, it
is important to add language on how introducing this
empricism could alter interpretation of our results. In particular,
while uWUE was developed using leaf scale photosythesis theory, it was
validated with observations of a fully coupled land-atmosphere
system. It captures a relatively consistent observed relationship
between GPP, ET, and VPD as they vary. There is a chance that some of
this relationship could be due to feedbacks between the land and the
atmosphere, even though the relationship was derived by leaf scale
theory without feedbacks (e.g., if the assumptions behind the uWUE
theory introduced ``real'' errors that were compensated for by
feedbacks or ecosystem-scale physics in the observations).  Were this
the case, then we introduced a conceptual impurity to the statement
``we are examining the one way response of ET to VPD.''  But, given
the theoretical motivations for uWUE we think the portion of the
relationship related to land-atmosphere feedbacks are likely small, if
they exist at all, and regardless this approach is still an
improvement over ignoring the documented effect of water stress on net
photosynthesis.

Given the continued gaps in our ability to represent stomatal behavior
\citep{Damour2010}, particularly for ecosystems in natural
environments, we think it is hard to rule out our results. In the end,
it comes down to the degree to which we trust our three tools: the
Medlyn model, Penman-Monteith, and uWUE. To us these offered the most
compete analytical approach to the problem at present, and are
supported in the literature. However, it's important to emphasize that
we are not saying this is certainly correct, and to do so would be
irresponsible given the continued development of stomatal theory and
how it translates to the ecosystem scale. What is important is to
recognize that something as fundamental as the shape of the ET-VPD
curve is still very much an area of open research, and sensitive to
the representation of plant physiology (of which there are many
possibilities) in a given model or framework. We thank the reviewer
for having an open mind and giving us a chance to elaborate; others in
the community have rejected this result off-hand, which reinforces for
us the importance of demonstrating current uncertainty in VPD-ET
relationships.

\vspace{0.2in}
Just a quick final note:

We think that we miscommunicated with our (mis)use of ``plant physiology''
throughout the manuscript, and will alter our language in the final
response phase to be more precise. Many times when we refer to ``plant
physiology'' we mean any effect (direct or indirect) of an ecosystem
response, e.g. any behavior that would not be observed over a wet or
bare soil surface.

\section{Major comment (iii)}
\label{iii}
Yes, through our manipulation of PM we have replaced stomatal
conductance with two parameters (g$_1$ and uWUE). However, given that
any physically reasonable representation of stomatal conductance will
include more parameters (usually at least some sort of VPD-related
parameter [g$_1$], and a model for photosynthesis with more
parameters), we think we are actually reducing the number of unknowns
in our model for ET. We agree with you that in reality g$_1$ and
uWUE are spatially variable, and we elaborate in detail in our
response to Reviewer 1
(\href{https://editor.copernicus.org/index.php/hess-2018-553-AC1.pdf?_mdl=msover_md&_jrl=13&_lcm=oc108lcm109w&_acm=get_comm_file&_ms=72556&c=153745&salt=1660920846993335212}{RC1
  Response})
on why and how it might be better to frame our results more in terms
of this variability. Also, see our comments in Section \ref{i} on the
effect of VPD on R$_n$ and the suitability of a binning approach to
VPD.

We also agree with you that uWUE includes a representation of ecosystem
response. However, we would argue that this response dependence of
stomatal conductance is not introduced by uWUE, but is actually an
issue that exists inherently in any stomatal conductance theory that
includes a photosynthesis term: photosynthesis depends on the
physiological response and environmental conditions. Using uWUE is our
(approximate) solution to this problem, and it works because we are
able to algebraically manipulate PM to isolate ET after introducing
uWUE. We do recognize that using uWUE introduces some uncertainty, but
as previously discussed in this response we believe it is a useful tool
for representing photosynthesis dependence of stomatal conductance in an
analytical framework.

\section{Major comment (iv)}

We agree on both counts: use of ``leading order'' is misguided and
confusing in this context. Also, better separation of theory and
empirical results would help communication/understanding.
\vspace{0.3in}

Thank you again for such a thoughtful review. Including these insights
will really improve the manuscript in the final response phase. In our
response to Reviewer 1
(\href{https://editor.copernicus.org/index.php/hess-2018-553-AC1.pdf?_mdl=msover_md&_jrl=13&_lcm=oc108lcm109w&_acm=get_comm_file&_ms=72556&c=153745&salt=1660920846993335212}{RC1
Response})
we also laid out some alternatives to our approach for the final
review phase not explicitly discussed here. If you (or anyone else)
has any opinions on that please weigh in.

\bibliography{references}
